---
title: "Untitled"
author: "Ludvig Larsson"
date: '2022-05-18'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r}
library(Seurat)
library(ggplot2)
```


Download files from Mendeley. You will need to download all spaceranger output folder to run this notebook which could take some time. 

You can skip this step if the data is already downloaded.

```{r eval=FALSE}

# Create data directories
suppressWarnings({
  dir.create("../data/")
  dir.create("../data/spaceranger_output/")
})

source("../../scripts/mendeley.R")

# Download spaceranger output files
for (i in seq_along(spaceranger.files)) {
  download.file(url = spaceranger.files[i], destfile = paste0("../data/spaceranger_output/", names(spaceranger.files)[i]))
  untar(tarfile = paste0("../data/spaceranger_output/", names(spaceranger.files)[i]), exdir = "../data/spaceranger_output/")
  file.remove(paste0("../data/spaceranger_output/", names(spaceranger.files)[i]))
}

```

Assemble spaceranger output files

```{r}

# Assemble output files
samples <- list.files(path = "../../data/spaceranger_output", pattern = "filtered_feature_bc_matrix", 
                      recursive = TRUE, full.names = TRUE)
imgs <- list.files(path = "../../data/spaceranger_output", pattern = "tissue_hires_image", 
                   recursive = TRUE, full.names = TRUE)
spotfiles <- list.files(path = "../../data/spaceranger_output", pattern = "tissue_positions_list", 
                       recursive = TRUE, full.names = TRUE)
json <- list.files(path = "../../data/spaceranger_output", pattern = "scalefactors_json", 
                   recursive = TRUE, full.names = TRUE)
metrics <- list.files(path = "../../data/spaceranger_output", pattern = "metrics_summary", 
                      recursive = TRUE, full.names = TRUE)

# Create infoTable
infoTable <- data.frame(samples, imgs, spotfiles, json)
infoTable <- cbind(infoTable, arrayID = do.call(rbind, strsplit(infoTable$samples, "/"))[, 6])

```

Merge metrics with additional metadata

```{r}

curated_metadata <- openxlsx::read.xlsx(xlsxFile = "../../data/sheets/RRST_sample_metadata.xlsx", 
                                        check.names = FALSE)
curated_metadata <- setNames(curated_metadata, nm = c("storage_time", "seq_date", "comments",
                                                       "ID", "paper_id", "RIN", "DV200", "protocol", "source", 
                                                       "arrayID", "spots_under_tissue",
                                                       "genes_detected", "fraction_spots_under_tissue",
                                                       "median_genes_per_spot", "median_UMIs_per_spot", 
                                                       "saturation", "reads_mapped_to_probe_set",
                                                       "reads_mapped_confidently_to_probe_set",
                                                       "reads_mapped_confidently_to_filtered_probe_set",
                                                       "reads_mapped_to_genome",
                                                       "reads_mapped_confidently_to_genome",
                                                       "number_of_panel_genes"))

infoTable_merged <- merge(infoTable, curated_metadata, by = "arrayID")

```

Calculate spot sample metrics

```{r warning=FALSE}

qc_all_samples <- do.call(rbind, lapply(seq_along(infoTable_merged$samples), function(i) {
  cat(sprintf("Loading sample %s\n", i))
  f <- infoTable_merged$samples[i]
  exprMat <- Seurat::Read10X_h5(f)
  data.frame(SpotID = colnames(exprMat), 
             UMIs = Matrix::colSums(exprMat), 
             unique_genes = Matrix::colSums(exprMat > 0),
             arrayID = infoTable_merged$arrayID[i])
}))

```

Here we abbreviate some of the longer IDs to fit in the plot. We also reorganize the samples in chronological order of the manuscript.

```{r}

qc_all_samples <- merge(qc_all_samples, infoTable_merged, by = "arrayID")
qc_all_samples$paper_id_ext <- qc_all_samples %>%
  mutate(paper_id_ext = case_when(paper_id == "P4" ~ "BNP4",
                                  paper_id == "P11" ~ "BNP11",
                                  paper_id == "medulloblastoma" ~ "PBT1",
                                  paper_id == "NOS" ~ "PBT2",
                                  TRUE ~ paper_id)) %>%
  pull(paper_id_ext)

# Sort levels
qc_all_samples$source <- factor(qc_all_samples$source, 
                                levels = c("mouse brain", "prostate cancer", "adult human lung", "adult human colon",
                                                                  "small intestine", "pediatric brain tumor", "mouse bone"))
qc_all_samples$paper_id_ext <- factor(qc_all_samples$paper_id_ext, 
                                      levels = c("MB", "PC", "LNG1", "LNG2", "CLN1", "CLN2", "SI", "PBT1", "PBT2", "BNP4", "BNP11"))

qc_all_samples$arrayID <- factor(qc_all_samples$arrayID, levels = qc_all_samples %>% 
                                   group_by(paper_id_ext, source, arrayID) %>% 
                                   tally() %>% pull(arrayID))

```

## Suppl. Figure 4 - Overview of unique genes distributions for all 52 samples
***

```{r fig.width=12, fig.height=5}

p <- ggplot(qc_all_samples, aes(arrayID, unique_genes, fill = protocol)) +
  geom_violin(scale = "width") +
  theme_bw() +
  theme(axis.title = element_blank(), 
        axis.text.y = element_text(colour = "black", size = 7),
        axis.text.x = element_text(angle = 60, hjust = 1, color = "black"), 
        strip.text.x = element_text(size = 8, color = "white", face = "bold"), 
        strip.text.y = element_text(size = 7, color = "white"), 
        strip.background = element_rect(fill = "#3A3B3C"),
        legend.position = "top") +
  labs(title = "Unique genes per spot") +
  facet_grid(paper_id_ext ~ source, scales = "free", space = "free", 
             labeller = labeller(source = label_wrap_gen(width = 12))) +
  scale_fill_manual(values = c("#4477AA", "#DDCC77"))

pdf(file = "QC_all_samples_unique_genes.pdf", width = 10, height = 12)
print(p)
dev.off()

png(filename = "QC_all_samples_unique_genes.png", width = 1000*2.5, height = 1200*2.5, res = 300)
print(p)
dev.off()

```

## Suppl. Figure 4 - Overview of UMI distributions for all 52 samples
***

```{r fig.width=12, fig.height=5}

p <- ggplot(qc_all_samples, aes(arrayID, UMIs, fill = protocol)) +
  geom_violin(scale = "width") +
  theme_bw() +
  theme(axis.title = element_blank(), 
        axis.text.y = element_text(colour = "black", size = 7),
        axis.text.x = element_text(angle = 60, hjust = 1, color = "black"), 
        strip.text.x = element_text(size = 8, color = "white", face = "bold"), 
        strip.text.y = element_text(size = 7, color = "white"), 
        strip.background = element_rect(fill = "#3A3B3C"),
        legend.position = "top") +
  labs(title = "UMIs per spot") +
  facet_grid(paper_id_ext ~ source, scales = "free", space = "free", 
             labeller = labeller(source = label_wrap_gen(width = 12))) +
  scale_fill_manual(values = c("#4477AA", "#DDCC77")) +
  scale_y_log10()

pdf(file = "QC_all_samples_UMIs.pdf", width = 10, height = 12)
print(p)
dev.off()

png(filename = "QC_all_samples_UMIs.png", width = 1000*2.5, height = 1200*2.5, res = 300)
print(p)
dev.off()

```


# Date
***

```{r}
date()
```

# Session
***

```{r}
devtools::session_info()
```


