---
title: "Figure 4"
author: "Ludvig Larsson"
date: '2022-08-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r eval=FALSE}

library(STutility)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(magrittr)

```
Download files from Mendeley.

```{r eval=FALSE}

# Create data directories
suppressWarnings({
  dir.create("../data/")
  dir.create("../data/genes/")
  dir.create("../data/sheets/")
  dir.create("../data/spaceranger_output/")
})

source("../scripts/mendeley.R")

# Download gene annotation files
for (i in seq_along(genes.files)) {
  download.file(url = genes.files[i], destfile = paste0("../data/genes/", names(genes.files)[i]))
}

# Download gene annotation files
for (i in seq_along(sheets.files)) {
  download.file(url = sheets.files[i], destfile = paste0("../data/sheets/", names(sheets.files)[i]))
}

# Download spaceranger output files
download.file(url = spaceranger.files[1], destfile = paste0("../data/spaceranger_output/", names(spaceranger.files)[1]))
untar(tarfile = paste0("../data/spaceranger_output/", names(spaceranger.files)[1]), exdir = "../data/spaceranger_output/")
file.remove(paste0("../data/spaceranger_output/", names(spaceranger.files)[1]))

```

Assemble spaceranger output files and merge curated meta data

```{r eval=FALSE}

# Mouse brain
samples <- Sys.glob(paths = "../../spaceranger_output/childbraintumor/*/filtered_feature_bc_matrix.h5")
imgs <- Sys.glob(paths = "../../spaceranger_output/childbraintumor/*/spatial/tissue_hires_image.png")
spotfiles <- Sys.glob(paths = "../../spaceranger_output/childbraintumor/*/spatial/tissue_positions_list.csv")
json <- Sys.glob(paths = "../../spaceranger_output/childbraintumor/*/spatial/scalefactors_json.json")

infoTable <- data.frame(samples, imgs, spotfiles, json)
infoTable <- cbind(infoTable, arrayID = do.call(rbind, strsplit(infoTable$samples, "/"))[, 5])

curated_metadata <- openxlsx::read.xlsx("../../sheets/curated_RNA_rescue_sample_metadata.xlsx", sheet = 3)
curated_metadata <- setNames(curated_metadata, nm = c("storage_time", "seq_date", "include", "RNA_rescue",
                                                       "project", "experimenter", "processer", "comments",
                                                       "ID", "paper_id", "RIN", "DV200", "protocol", "source", "arrayID", "spots_under_tissue",
                                                       "genes_detected", "fraction_spots_under_tissue",
                                                       "median_genes_per_spot", "median_UMIs_per_spot", 
                                                       "saturation", "reads_mapped_to_probe_set",
                                                       "reads_mapped_confidently_to_probe_set",
                                                       "reads_mapped_confidently_to_filtered_probe_set",
                                                       "reads_mapped_to_genome",
                                                       "reads_mapped_confidently_to_genome",
                                                       "number_of_panel_genes"))

infoTable <- merge(infoTable, curated_metadata, by = "arrayID")

```

Load data into a Seurat object


```{r eval=FALSE}

CBT <- InputFromTable(infoTable)

```

```{r eval=FALSE, fig.width=12, fig.height=8}

CBT$protocol_array <- paste0(CBT$protocol, " : ", CBT$arrayID)
ST.FeaturePlot(CBT, features = "nFeature_RNA", ncol = 4, label.by = "protocol_array", show.sb = FALSE, pt.size = 1.5)

```

Load images

```{r eval=FALSE}

CBT <- LoadImages(CBT, time.resolve = FALSE, xdim = 1e3)

```

```{r}

ImagePlot(CBT, method = "raster", ncols = 4, type = "raw")

```


Add empty masks for downstream processing of HE images

```{r eval=FALSE}

# Prepare data for rigid transformation
st.object <- GetStaffli(CBT)
msks <- lapply(st.object@rasterlists$raw, function(im) {
  as.raster(matrix(data = 1, ncol = ncol(im), nrow = nrow(im)))
})
st.object@rasterlists$masked <- st.object@rasterlists$raw
st.object@rasterlists$masked.masks <- msks
CBT@tools$Staffli <- st.object

```


Apply transformations

```{r eval=FALSE}

# Warp transform
CBT <- WarpImages(CBT, verbose = TRUE, transforms = list("2" = list(shift.y = 40*2.5),
                                                         "3" = list(shift.y = 40*2.5),
                                                         "4" = list(shift.y = 40*2.5),
                                                         "6" = list(shift.y = 20*2.5),
                                                         "7" = list(shift.y = 30*2.5, shift.x = -30*2.5),
                                                         "8" = list(shift.x = 20*2.5)))

```

```{r}

p_violin <- ggplot() + 
  geom_violin(data = CBT[[]], 
              aes(arrayID, nFeature_RNA, fill = protocol), scale = "width") + 
  geom_hline(data = CBT[[]] %>% group_by(protocol, paper_id) %>%
               summarize(Mean = round(mean(nFeature_RNA))), 
             aes(yintercept = Mean), linetype = "longdash") +
  geom_label(data = CBT[[]] %>% group_by(protocol, paper_id) %>%
               summarize(Mean = round(mean(nFeature_RNA))), 
             aes("A", Mean, label = Mean), size = 7) +
  facet_grid(~paper_id, scales = "free", space = "free", 
             labeller = labeller(paper_id = c("1.0" = "medulloblastoma", "2.0" = "NOS"))) +
  scale_x_discrete(labels = c("", "Rep1", "Rep2", "Rep1", "Rep2")) +
  scale_y_log10() +
  #scale_fill_brewer(palette = "Pastel2") +
  labs(x = "", y = "") +
  theme(panel.background = element_rect(fill = "white"), panel.grid = element_line(colour = "lightgray", size = 0.5), 
        strip.text = element_text(size = 16, colour = "black"), 
        axis.text = element_text(size = 16, colour = "black"), 
        legend.text = element_text(size = 14, colour = "black"),
        legend.title = element_text(size = 16, colour = "black"), 
        legend.position = "bottom")

pdf(file = "unique_genes_overview.pdf", width = 8, height = 6)
print(p_violin)
dev.off()

```

Spatial unique genes plots

```{r}

# Spatial plots showing the number of unique genes
df_subset <- cbind(CBT[[]], GetStaffli(CBT)@meta.data) %>%
  subset(sample %in% c("1", "3", "5", "7"))
scale_limits <- c(0, max(df_subset$nFeature_RNA))
x_limits <- range(df_subset$warped_x)
y_limits <- range(df_subset$warped_y)
p_spatial <- df_subset %>%
    ggplot(aes(warped_x, 2000 - warped_y, color = nFeature_RNA)) +
      geom_point(size = 1) +
      facet_wrap(paper_id ~ protocol, labeller = labeller(paper_id = c("1.0" = "CBT1", "2.0" = "CBT2")), ncol = 4) +
      theme_void() +
      labs(color = "Unique\ngenes") +
      scale_color_gradientn(colours = viridis::magma(n = 11, direction = -1), limits = scale_limits) +
      scale_x_continuous(limits = x_limits, expand = c(0, 0)) +
      scale_y_continuous(limits = c(2000 - y_limits[2], 2000 - y_limits[1]), expand = c(0, 0)) +
      theme(strip.text = element_blank(), 
            panel.background = element_rect(colour = "lightgray", size = 2),
            legend.text = element_text(size = 14, colour = "black"),
            legend.title = element_text(size = 16, colour = "black"))
p_spatial <- ggrastr::rasterize(p_spatial, layers = "Point", dpi = 300)

# HE images
plots_HE <- lapply(c("1", "3", "5", "7"), function(i) {
  im <- GetStaffli(CBT)@rasterlists$processed[[i]]
   
  im <- im[round(y_limits[1]*0.5):round( y_limits[2]*0.5), round(x_limits[1]*0.5):round(x_limits[2]*0.5)]
  g <- grid::rasterGrob(im, width = unit(1, "npc"), height = unit(1, "npc"), interpolate = TRUE)
  p <- ggplot() +
    annotation_custom(g, -Inf, Inf, -Inf, Inf) +
    theme_void() +
    theme(plot.margin = margin(t = 1, r = 5, b = 1, l = 5))
  return(p)
})

```

Create patchwork for HE and spatial visualization of unique genes per spot

```{r}

design <- c(
  patchwork::area(1, 1, 1, 1),
  patchwork::area(1, 2, 1, 2),
  patchwork::area(1, 3, 1, 3),
  patchwork::area(1, 4, 1, 4),
  patchwork::area(2, 1, 2, 4)
)

# Create patchwork
p_HE <- plots_HE[[3]] +
  plots_HE[[1]] +
  plots_HE[[4]] +
  plots_HE[[2]] +
  p_spatial +
  patchwork::plot_layout(design = design)

pdf(file = "HE_overivew.pdf", width = 14, height = 6)
print(p_HE)
dev.off()

```

Read annotations

```{r}

annotations <- Sys.glob(paths = "../../spaceranger_output/childbraintumor/*/Pathology*")

ann <- do.call(rbind, lapply(seq_along(annotations), function(i) {
  a <- setNames(read.csv(file = annotations[i]), nm = c("barcode", "type"))
  a <- data.frame(a, row.names = 1)
  rownames(a) <- paste0(rownames(a), "_", i + 4)
  return(a)
}))

CBT$annotation <- ann[colnames(CBT), "type"]
CBT$annotation[CBT$annotation == "N/A"] <- "unassigned"

```

Plot annotations

```{r fig.width=5, fig.height=5}

p <- FeatureOverlay(CBT, features = "annotation", sampleids = 5, show.sb = FALSE) &
  theme(plot.title = element_blank(), plot.subtitle = element_blank())

im <- GetStaffli(CBT)@rasterlists$processed[[5]]
im <- im[round(500*0.5):round( y_limits[2]*0.5), round(x_limits[1]*0.5):round(x_limits[2]*0.5)]
g <- grid::rasterGrob(im, width = unit(1, "npc"), height = unit(1, "npc"), interpolate = TRUE)
p <- ggplot() +
  annotation_custom(g, -Inf, Inf, -Inf, Inf) +
  geom_point(data = cbind(CBT[[]], GetStaffli(CBT)@meta.data) %>% subset(sample == "5"), 
             aes(warped_x, 2000 - warped_y, color = annotation), alpha = 1) +
  scale_x_continuous(limits = x_limits, expand = c(0, 0)) +
  scale_y_continuous(limits = c(2000 - y_limits[2], 2000 - 500), expand = c(0, 0)) +
  scale_color_manual(values = c("tumor" = "#117733", "stroma" = "#DDCC77", "unassigned" = "lightgrey")) +
  theme_void() +
  guides(colour = guide_legend(override.aes = list(size = 5))) +
  theme(plot.margin = margin(t = 1, r = 5, b = 1, l = 5), 
        legend.position = "bottom", 
        legend.text = element_text(size = 14, colour = "black"),
        legend.title = element_blank())
p <- ggrastr::rasterize(p, layers = "Point", dpi = 300)

pdf(file = "medulloblastoma_annotations.pdf", width = 5, height = 4.2)
print(p)
dev.off()


```


Subset data to include medulloblastoma only

```{r fig.width=8, fig.height=3}

CBT_medulloblastoma <- SubsetSTData(CBT, expression = paper_id == "1.0")
CBT_medulloblastoma <- CBT_medulloblastoma %>%
  NormalizeData() %>%
  ScaleData() %>%
  FindVariableFeatures()
ST.FeaturePlot(CBT_medulloblastoma, features = "AXIN2", ncol = 4, show.sb = FALSE)

```

Visualize marker genes

```{r fig.width=12, fig.height=7}

WNT_markers <- c("SP5", "AXIN2", "LEF1", "DKK4", "GAD1", "CTNNB1")

p_markers <- cbind(CBT_medulloblastoma[[]], FetchData(CBT_medulloblastoma, vars = WNT_markers)) %>%
  reshape2::melt(measure.vars = WNT_markers) %>%
  ggplot(aes(arrayID, value, fill = protocol)) +
    geom_violin(scale = "width") +
    scale_x_discrete(labels = c("Rep1", "Rep2", "Rep1", "Rep2")) +
    facet_wrap(~variable, ncol = 3) +
    theme(panel.background = element_rect(fill = "white"),
          panel.grid = element_line(linetype = "longdash", colour = "lightgrey"),
          strip.text = element_text(size = 14, colour = "black"),
          axis.text.x = element_text(size = 12, colour = "black"),
          axis.text.y = element_text(size = 14, colour = "black"),
          axis.title = element_text(size = 16, colour = "black")) +
  #scale_fill_brewer(palette = "Pastel2") +
  labs(y = "Norm. Expr.", x = "")
p_markers

pdf(file = "violin_markers.pdf", width = 8, height = 5)
print(p_markers)
dev.off()

```


Visualize pathway activities

```{r}

read.gmt <- function(filename) {
  d <- suppressWarnings({(readLines(filename) %>% strsplit("\t"))[[1]]})
  setNames(list(d[3:length(d)]), nm = d[1])
}

WNT_genes <- read.gmt(filename = "pathway_gmt_files/HALLMARK_WNT_BETA_CATENIN_SIGNALING.v7.5.1.gmt")
TGFB_genes <- read.gmt(filename = "pathway_gmt_files/HALLMARK_TGF_BETA_SIGNALING.v7.5.1.gmt")

```

```{r fig.width=5, fig.height=5}

CBT_medulloblastoma <- AddModuleScore(CBT_medulloblastoma, features = c(WNT_genes, TGFB_genes), name = "pathway")

p_pathway <- cbind(CBT_medulloblastoma[[]], GetStaffli(CBT_medulloblastoma)@meta.data) %>%
  subset(sample %in% c("1", "3")) %>%
  reshape2::melt(measure.vars = c("pathway2", "pathway1")) %>%
  mutate(value = ifelse(value > quantile(value, 0.99), quantile(value, 0.99), value)) %>%
  ggplot(aes(warped_x, 2000 - warped_y, color = value)) +
    geom_point(size = 1) +
    facet_wrap(variable ~ protocol) +
    scale_color_gradientn(colours = RColorBrewer::brewer.pal(n = 11, name = "Spectral") %>% rev()) +
    theme_void() +
    theme(strip.text = element_blank(), panel.background = element_rect(colour = "lightgrey", size = 2),
          legend.text = element_text(size = 14, colour = "black"),
          legend.title = element_text(size = 16, colour = "black"))
p_pathway <- ggrastr::rasterize(p_pathway, layers = "Point", dpi = 300)

pdf(file = "pathway_spatial_plot.pdf", width = 8, height = 6.5)
print(p_pathway)
dev.off()

```


