---
title: "Figure 4"
author: "Ludvig Larsson"
date: '2022-07-04'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries

```{r eval=FALSE}

library(STutility)
library(ggplot2)
library(ggpubr)

```

Assemble spaceranger output files and merge curated meta data

```{r eval=FALSE}

# Mouse brain
samples <- Sys.glob(paths = "../../spaceranger_output/smallintestine/*/filtered_feature_bc_matrix.h5")
imgs <- Sys.glob(paths = "../../spaceranger_output/smallintestine/*/spatial/tissue_hires_image.png")
spotfiles <- Sys.glob(paths = "../../spaceranger_output/smallintestine/*/spatial/tissue_positions_list.csv")
json <- Sys.glob(paths = "../../spaceranger_output/smallintestine/*/spatial/scalefactors_json.json")

infoTable <- data.frame(samples, imgs, spotfiles, json)
infoTable <- cbind(infoTable, arrayID = do.call(rbind, strsplit(infoTable$samples, "/"))[, 5])

curated_metadata <- openxlsx::read.xlsx("../../sheets/curated_RNA_rescue_sample_metadata.xlsx", sheet = 3)
curated_metadata <- setNames(curated_metadata, nm = c("storage_time", "seq_date", "include", "RNA_rescue",
                                                       "project", "experimenter", "processer", "comments",
                                                       "ID", "paper_id", "protocol", "source", "arrayID", "spots_under_tissue",
                                                       "genes_detected", "fraction_spots_under_tissue",
                                                       "median_genes_per_spot", "median_UMIs_per_spot", 
                                                       "saturation", "reads_mapped_to_probe_set",
                                                       "reads_mapped_confidently_to_probe_set",
                                                       "reads_mapped_confidently_to_filtered_probe_set",
                                                       "reads_mapped_to_genome",
                                                       "reads_mapped_confidently_to_genome",
                                                       "number_of_panel_genes"))

infoTable <- merge(infoTable, curated_metadata, by = "arrayID")

```

Load data into a Seurat object


```{r eval=FALSE}

SI <- InputFromTable(infoTable %>% arrange(seq_date))

```

```{r eval=FALSE, fig.width=10, fig.height=12}

SI$protocol_array <- paste0(SI$protocol, " : ", SI$arrayID)
ST.FeaturePlot(SI, features = "nFeature_RNA", ncol = 3, label.by = "protocol_array", show.sb = FALSE, pt.size = 1.5)

```

Load images

```{r eval=FALSE}

SI <- LoadImages(SI, time.resolve = FALSE, xdim = 1e3)

```

Add empty masks for downstream processing of HE images

```{r eval=FALSE}

# Prepare data for rigid transformation
st.object <- GetStaffli(SI)
msks <- lapply(st.object@rasterlists$raw, function(im) {
  as.raster(matrix(data = 1, ncol = ncol(im), nrow = nrow(im)))
})
st.object@rasterlists$masked <- st.object@rasterlists$raw
st.object@rasterlists$masked.masks <- msks
SI@tools$Staffli <- st.object

```

Apply transformations

```{r eval=FALSE}

# Warp transform
SI <- WarpImages(SI, verbose = TRUE, transforms = list("3" = list(angle = -90), "4" = list(angle = -90), 
                                                         "5" = list(angle = 160), "6" = list(angle = 160),
                                                         "7" = list(angle = 20), "8" = list(angle = 20),
                                                         "9" = list(angle = 20), "10" = list(angle = 20),
                                                         "11" = list(angle = 20), "12" = list(angle = 20),
                                                         "13" = list(angle = -140), "14" = list(angle = 30)))

```

Add a new metadata column with a combined protocol and lung ID label

```{r eval=FALSE, fig.width=10, fig.height=10}

SI$protocol_lng <- gsub(pattern = " ", replacement = "_", paste0(SI$protocol, "_ID: ", SI$paper_id))
ST.FeaturePlot(SI, features = "nFeature_RNA", ncol = 3, label.by = "protocol_lng")

```

Add manual annotations

```{r eval=FALSE}

SI <- ManualAnnotation(SI, type = "raw")

```

```{r include=FALSE, eval=FALSE}
# Export selections

#saveRDS(SI@meta.data, file = "../R_objects/SI_metadata_selections")
meta.data <- readRDS("../R_objects/SI_metadata_selections")
meta.data <- meta.data[colnames(SI), ]
SI@meta.data$labels <- meta.data$labels
SI@meta.data$labels[is.na(SI@meta.data$labels)] <- "background"

submucosa_13 <- rownames(subset(SI@meta.data, arrayID == "V10S29-108_A1" & labels == "background"))
SI@meta.data[submucosa_13, "labels"] <- "submucosa"

```

```{r fig.width=12, fig.height=10}

ST.FeaturePlot(SI, features = "labels", ncol = 4)

```



### Filter out background and serosa
***

```{r eval=FALSE}

SI <- SubsetSTData(SI, expression = labels %in% c("mucosa", "submucosa", "muscularis", "TLS", "serosa"))

```


### Biotype content
***

```{r}

ensids <- read.table("../../genes/hgenes.tsv", header = T)
rownames(ensids) <- ensids$gene_name

ensids$gene_type[grep(pattern = "^MT-", x = ensids$gene_name)] <- "mt_protein_coding"
ensids$gene_type[grep(pattern = "^RPL|^RPS", x = ensids$gene_name)] <- "rb_protein_coding"

```


```{r}

SI$time <- SI[[]] %>%
  mutate(time = factor(case_when(seq_date == "200323" ~ "~ 1 month",
                          seq_date == "200923" ~ "~ 6 months",
                          seq_date %in% c("220404", "220506") ~ "~ 2 years"), 
                       levels = c("~ 1 month", "~ 6 months", "~ 2 years"))) %>% # ifelse(SI$seq_date %in% c("220404", "220506"), "220404-220506", SI$seq_date)
  pull(time)

gene_attr <- do.call(rbind, lapply(levels(SI$time), function(i) {
  x <- rowSums(GetAssayData(SI, slot = "counts", assay = "RNA")[, SI$time == i])
  data.frame(gene = names(x), count = x, gene_type = ensids[names(x), "gene_type"], time = i)
}))

gene_attr <- gene_attr %>% group_by(time, gene_type) %>% summarize(N = sum(count)) %>% group_by(time) %>% mutate(p = N/sum(N)) %>%
  mutate(gene_type = factor(ifelse(gene_type %in% c("IG_C_gene", "IG_J_gene", "IG_V_gene",
                          "TR_C_gene", "TR_J_gene", "TR_V_gene",
                          "lncRNA", "protein_coding", "rb_protein_coding",
                          "mt_protein_coding"), gene_type, "other"), levels = 
                            c("protein_coding", "rb_protein_coding",
                          "mt_protein_coding", "lncRNA", "IG_C_gene", "IG_J_gene", "IG_V_gene",
                          "TR_C_gene", "TR_J_gene", "TR_V_gene", "other")))
gene_attr$time <- factor(gene_attr$time, levels = c("~ 1 month", "~ 6 months", "~ 2 years"))

cols <- c("#332288", "#6699CC", "#88CCEE", "#44AA99", "#117733", "#999933", "#DDCC77", "#661100", "#CC6677", "#882255", "#AA4499")
p <- ggplot(gene_attr, aes(x = "", y = p, fill = gene_type)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  facet_wrap(~ time, ncol = 2) +
  theme(panel.background = element_rect(fill = "white"), 
        axis.ticks = element_blank(), axis.text.x = element_blank(),
        strip.text = element_text(size = 14, colour = "black"), 
        legend.text = element_text(size = 14, colour = "black"), 
        legend.position = c(0.8, 0.3)) +
  labs(x = "", y = "", fill = "") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = cols)

pdf(file = "biotype_content.pdf", width = 6, height = 6)
print(p)
dev.off()

```

### Plot H&E
***

```{r}

png(filename = "HE.png", width = 860 - 95, height = 883 - 116)
#95, 116, 860, 883
par(mar = c(0,0,0,0))
plot(SI@tools$Staffli@rasterlists$raw$`2`[116:883, 95:860])
dev.off()

```

### Plot annotation
***

```{r}
  
gg <- cbind(SI[[]], GetStaffli(SI)@meta.data)
gg <- subset(gg, sample == "2")
dims <- GetStaffli(SI)@dims[["2"]]

limits_list <- list("2" = c(x_start = 95*2, y_start = 116*2, x_end = 860*2, y_end = 883*2))

p <- ggplot(gg, aes(warped_x, dims$height - warped_y, color = labels)) +
  geom_point(size = 1.7) +
  theme_void() +
  scale_color_manual(values = c("mucosa" = "#AA4499", "submucosa" = "#DDCC77", "muscularis" = "#CC6677", "TLS" = "#77AADD", "serosa" = "#771155")) +
  scale_x_continuous(expand = c(0, 0), limits = c(limits_list[["2"]]["x_start"], 
                                limits_list[["2"]]["x_end"])) +
  scale_y_continuous(expand = c(0, 0), limits = c(dims$height - limits_list[["2"]]["y_end"], 
                                dims$height - limits_list[["2"]]["y_start"])) +
  theme(legend.position = "none")

p <- ggrastr::rasterize(p, layers = "Point", dpi = 300)


png(filename = "annotations.png", width = (860 - 95)*2, height = (883 - 116)*2, res = 300)
print(p)
dev.off()

```


### Add new date column
***

```{r fig.width=12, fig.height=8}
  
p <- ggplot() + 
  geom_violin(data = SI[[]], 
              aes(arrayID, nFeature_RNA, fill = protocol), scale = "width") + 
  geom_hline(data = SI[[]] %>% group_by(protocol, time, labels) %>%
               summarize(Mean = round(mean(nFeature_RNA))), 
             aes(yintercept = Mean), linetype = "longdash") +
  geom_label(data = SI[[]] %>% group_by(protocol, time, labels) %>%
               summarize(Mean = round(mean(nFeature_RNA))), 
             aes("A", Mean, label = Mean), size = 5) +
  facet_grid(labels ~ time, scales = "free", space = "free") +
  scale_x_discrete(labels = c("", "Rep1", "Rep2", "Rep3", "Rep4", "Rep5", "Rep6", "Rep7", "Rep8")) +
  scale_y_log10() +
  scale_fill_brewer(palette = "Pastel2") +
  labs(x = "", y = "") +
  theme(panel.background = element_rect(fill = "white"), panel.grid = element_line(colour = "lightgray", size = 0.5), 
        strip.text = element_text(size = 14, colour = "black"), 
        axis.text = element_text(size = 14, colour = "black"), 
        legend.text = element_text(size = 14, colour = "black"),
        legend.title = element_text(size = 16, colour = "black"), 
        legend.position = "bottom")

pdf(file = "unique_genes_overview.pdf", width = 14, height = 10)
print(p)
dev.off()

```


### Export annotations (Suppl?)
***

```{r fig.width=14, fig.height=4}


infoTable[, 1]

limits_list <- list("1" = c(x_start = 260, y_start = 500, x_end = 1600, y_end = 1800),
                    "5" = c(x_start = 170, y_start = 340, x_end = 1400, y_end = 1500),
                    "13" = c(x_start = 130, y_start = 240, x_end = 1780, y_end = 1560))

ann_plots <- lapply(c("1", "5", "13"), function(i) {
  
  gg <- cbind(SI[[]], GetStaffli(SI)@meta.data)
  gg <- subset(gg, sample == i)
  dims <- GetStaffli(SI)@dims[[i]]
  
  p <- ggplot(gg, aes(warped_x, dims$height - warped_y, color = labels)) +
    geom_point(size = 1) +
    theme_void() +
    scale_color_manual(values = c("mucosa" = "#AA4499", "submucosa" = "#DDCC77", "muscularis" = "#CC6677", 
                                  "TLS" = "#77AADD", "serosa" = "#771155")) +
    scale_x_continuous(expand = c(0, 0), limits = c(limits_list[[i]]["x_start"], 
                                  limits_list[[i]]["x_end"])) +
    scale_y_continuous(expand = c(0, 0), limits = c(dims$height - limits_list[[i]]["y_end"], 
                                  dims$height - limits_list[[i]]["y_start"])) +
    theme(legend.position = "none")

  p <- ggrastr::rasterize(p, layers = "Point", dpi = 300)
  return(p)
})

# HE images
plots_HE <- lapply(c("2", "3", "4", "6"), function(i) {
  im <- GetStaffli(CLN)@rasterlists$processed[[paste0(i)]]
  dims <- GetStaffli(CLN)@dims[[i]]
   
  im <- im[limits_list[[i]]["y_start"]:limits_list[[i]]["y_end"]*0.5, limits_list[[i]]["x_start"]:limits_list[[i]]["x_end"]*0.5]
  g <- grid::rasterGrob(im, width = unit(1, "npc"), height = unit(1, "npc"), interpolate = TRUE)
  p <- ggplot() +
    annotation_custom(g, -Inf, Inf, -Inf, Inf) +
    theme_void() +
    theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0))
  return(p)
})


```


# Analysis
***

```{r}

SISplit <- lapply(levels(SI$time), function(i) {
  cat(sprintf("Started processing data set %s", i), "\n")
  SIs <- SubsetSTData(SI, expression = nFeature_RNA > 100)
  SIs <- SubsetSTData(SIs, expression = time %in% i)
  SIs <- SIs %>% 
    NormalizeData() %>% 
    ScaleData() %>% 
    FindVariableFeatures() %>%
    RunPCA() %>%
    RunHarmony(group.by.vars = "arrayID", reduction = "pca", dims.use = 1:20) %>%
    FindNeighbors(reduction = "harmony", dims = 1:20) %>%
    FindClusters() %>%
    RunUMAP(reduction = "harmony", dims = 1:20)
  SIs <- SetIdent(SIs, value = "labels")
  return(SIs)
})

```

```{r fig.width=6, fig.height=5}

DimPlot(SISplit[[3]], group.by = "seurat_clusters") +
DimPlot(SISplit[[3]], group.by = "arrayID") +
  DimPlot(SISplit[[3]]) +
  patchwork::plot_layout(nrow = 2)

```


DE test

```{r}

de.markers.list <- lapply(SISplit, function(se) {
  de.markers <- FindMarkers(se, only.pos = TRUE, max.cells.per.ident = 1000, ident.1 = "mucosa") %>% filter(p_val_adj < 0.01)
  de.markers$gene <- rownames(de.markers)
  de.markers$time <- unique(se$time)
  return(de.markers)
})

```



```{r}

de.markers <- do.call(rbind, de.markers.list)

expressionInput <- de.markers %>% group_by(gene) %>%
  summarize(group = paste0(time, collapse = "&")) %>%
  group_by(group) %>%
  summarize(N = n()) %>%
  arrange(-N)
expressionInput <-  setNames(expressionInput$N, nm =  expressionInput$group)

p <- upset(fromExpression(expressionInput), order.by = "freq")

pdf(file = "upset_DE.pdf", width = 6, height = 3)
print(p)
dev.off()

```

```{r}

SI <- NormalizeData(SI)

enterocyte.markers <- c("ANPEP", "RBP2", "DGAT1", "FABP2", "APOB")
gg <- cbind(SI[[]], GetStaffli(SI)@meta.data, FetchData(object = SI, vars = enterocyte.markers)) %>%
  reshape::melt(measure.vars = enterocyte.markers) %>%
  select(arrayID, time, warped_x, warped_y, sample, variable, value)

plots <- lapply(enterocyte.markers, function(i) {
  subset(gg, variable == i) %>%
  subset(sample %in% c("2", "5", "13")) %>%
    ggplot(aes(warped_x, warped_y, color = value)) +
      geom_point(size = 0.3) +
      theme_void() +
      facet_grid(time ~ .) +
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_reverse(expand = c(0, 0)) +
      scale_color_gradientn(colours = viridis::magma(n = 11, direction = -1)) +
      labs(title = i) +
      theme(legend.position = "top", 
            strip.text = element_blank()) -> p
  ggrastr::rasterize(p, layers = "Point", dpi = 300)
})

p <- cowplot::plot_grid(plotlist = plots, nrow = 1)

pdf(file = "markes_spatial_enterocytes.pdf", width = 12, height = 8)
print(p)
dev.off()

```
Export H&E images

```{r}

for (i in c("2", "5", "13")) {
  im <- SI@tools$Staffli@rasterlists$processed[[i]]
  png(filename = paste0("HE_", i, ".png"), width = ncol(im), height = nrow(im), res = 300)
  par(mar = c(0,0,0,0))
  plot(im)
  dev.off()
}

```


```{r}

de.markers %>% mutate(Freq = as.numeric(table(gene)[gene])) %>%
  filter(Freq == 3) %>%
  group_by(gene) %>%
  summarize(Mean = mean(avg_log2FC)) %>%
  arrange(-Mean)

```

Scale data

```{r fig.width=6, fig.height=2}

SI <- NormalizeData(SI)
SI <- ScaleData(SI, features = de.markers$gene[1:10])
DotPlot(SI, features = de.markers$gene[1:10])

gg <- cbind(SI[[]], FetchData(SI, vars = de.markers$gene[1:10], slot = "data")) %>% 
  reshape2::melt(measure.vars = de.markers$gene[1:10]) %>%
  group_by(variable, labels) %>%
  summarize(Mean = mean(value), pct = sum(value > 0)/n())

ggplot(gg, aes(variable, labels, color = Mean, size = pct)) +
  geom_point() +
  #facet_grid(~time) +
  scale_color_gradientn(colours = viridis::magma(n = 11, direction = -1))

```

DE test by region

```{r fig.width=8, fig.height=3}

region.markers <- FindAllMarkers(SI, only.pos = TRUE)
selected.markers <- region.markers %>% group_by(cluster) %>% slice_head(n = 10) %>% pull(gene)

SI <- ScaleData(SI, features = selected.markers)
DotPlot(SI, features = selected.markers, split.by = "time", cols = c("red", "orange", "blue"), dot.scale = 7)

```


Gene ontology

```{r}

SI$mucosa_bg <- ifelse(SI$labels == "mucosa", "mucosa", "bg")

ST.FeaturePlot(SI, features = "mucosa_bg", indices = c(2, 9, 13), cols = c("mucosa" = "red", "bg" = "lightgray")) &
  theme(strip.text = element_blank())

```



```{r}

DotPlot(SI, features = de.markers$gene[1:10])

```






# date
***

```{r eval=FALSE}
date()
```

# Session
***

```{r eval=FALSE}
devtools::session_info()
```